Supporting information - Bayesian regression facilitates quantitative modelling of cell metabolism

### 2024-01-12 paper/main.qmd
review response changes

Maud represents information from measurements using generalised linear
regression models that probabilistically connect realised measurements with
true values of the measurable quantities. Information from other sources
is represented using a prior distribution over a set of latent parameters.
The parameters and the measurable quantities are connected by a generative
model encompassing Maud’s kinetic model as well as the steady state equation
\eqref{eq-3}. Together, the measurement model, prior model and generative model
determine a joint probability function that assigns a probability density to any
possible combination of measurements and parameters.

Below we describe the prior and measurement models, as the generative model has
already been discussed above.

>>>>

Maud uses Markov Chain Monte Carlo to perform posterior inference on a Bayesian
statistical model. This section introduces these topics and then describes
exactly what kind of Bayesian statistical model Maud uses.

### Bayesian inference

Bayesian statistical inference analyses data by constructing a mathematical
model with the following three components:

- A measurement model or "likelihood" that probabilistically describes
how likely any possible measurement set is given the true values
of the measured quantities, i.e. a probability density function $l:
\mathfrak{Y}\times\hat{\mathfrak{Y}} \rightarrow [0, 1]$ where $\mathfrak{Y}$
is the set of all possible measurement sets and $\hat{\mathfrak{Y}}$ is the set
of all possible true measurable values.

- A deterministic process model that describes the measured quantities in
terms of unknown, possibly unobserved parameters $\theta$, i.e. a function $d:
\mathfrak{theta} -> \hat{mathfrak{Y}}$.

- A prior model that probabilistically describes how likely any possible set
of parameter values is, without considering any information included in the
measurement model, i.e. a probability density function $\pi : \mathfrak{theta}
\rightarrow [0, 1]$.

Together these components determine a joint probability function
that encapsulates the Bayesian statistical model, i.e. a function $p:
\mathfrak{Y}\times\mathfrak{theta}\rightarrow[0,1]$ such that for any $\theta$
and $y$, $p(\theta, y) = \pi(\theta)l(y\mid\theta)$. Substantive questions about
the implications of the assumptions implicit in the model components can be
formulated in terms of this joint density function.

In particular, questions about what the model implies, given a particular
measurement set $y$ can be formulated in terms of the conditional probability
density function $p_{\mid y}:\mathfrak{theta}\rightarrow[0,1]$. Bayes's
theorem guarantees that this conditional density is proportional to the
product of the prior and likelihood, i.e. $p_{\mid y}(\theta) \propto
\pi(\theta)l(y\mid\theta)$ See [@gelmanBayesianDataAnalysis2020a, Ch 1] for an
extended discussion of the theory of Bayesian statistical inference.

The principal computational challenge for Bayesian statistical inference is to
approximate the values of integrals of the joint density function conditional
on a known measurement assignment $y$, also known as the posterior distribution.
For most non-trivial statistical models the posterior distribution cannot be
integrated analytically, so numerical approximation is required. Markov Chain
Monte Carlo or "MCMC" is a popular method for addressing this problem that
uses a Markov chain with known properties to generate samples from a target
probability distribution that can be used to perform Monte Carlo integration.
Hamiltonian Monte Carlo makes it possible, given an appropriate choice of
hyperparameters, to efficiently generate samples from even a high dimensional
continuous posterior distribution, by calculating the gradients of the log-scale
joint density function. Adaptive Hamiltonian Monte Carlo, as used by Stan, uses
a range of well-tested methods to tune these hyperparameters, allowing efficient
MCMC for high dimensional posterior distributions with minimal user input.

Alternatives to MCMC for numerically approximating integrals over posterior
distributions include grid sampling [@gelmanBayesianDataAnalysis2020a,
Ch. 10], rejection sampling [@gelmanBayesianDataAnalysis2020a, Ch. 10],
importance sampling [@vehtariParetoSmoothedImportance2022] and distributional
approaches including variational inference and Laplace approximation
[@gelmanBayesianDataAnalysis2020a, Ch. 13].

The Laplace approximation is of particular interest because, as mentioned above,
this has been used for approximate Bayesian inference in a similar context
to Maud. Laplace approximation of a posterior distribution works by first
finding the mode of the posterior distribution, i.e. the 'maximum a posteriori'
parameter configuration with the highest posterior density. Next, the second
derivative of the posterior distribution at this point is found and used to
generate a normal distribution that either approximates the target distribution,
or can in turn be used to generate such an approximation.

Maud employs adaptive Hamiltonian Monte Carlo to perform posterior inference
for a Bayesian statistical model where the process model is the kinetic model
described above. The next two subsections describe Maud's prior and measurement
models.

### 2024-01-12 paper/main.tex
review response changes

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

@@

Maud represents information from measurements using generalised linear
regression models that probabilistically connect realised measurements
with true values of the measurable quantities. Information from other
sources is represented using a prior distribution over a set of latent
parameters. The parameters and the measurable quantities are connected
by a generative model encompassing Maud's kinetic model as well as the
steady state equation \eqref{eq-3}. Together, the measurement model,
prior model and generative model determine a joint probability function
that assigns a probability density to any possible combination of
measurements and parameters.

Below we describe the prior and measurement models, as the generative
model has already been discussed above.

>>>>

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, interior hidden, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

Maud uses Markov Chain Monte Carlo to perform posterior inference on a
Bayesian statistical model. This section introduces these topics and
then describes exactly what kind of Bayesian statistical model Maud
uses.

\hypertarget{bayesian-inference}{%
\subsubsection{Bayesian inference}\label{bayesian-inference}}

Bayesian statistical inference analyses data by constructing a
mathematical model with the following three components:

\begin{itemize}
\item
  A measurement model or ``likelihood'' that probabilistically describes
  how likely any possible measurement set is given the true values of
  the measured quantities, i.e.~a probability density function
  \(l: \mathfrak{Y}\times\hat{\mathfrak{Y}} \rightarrow [0, 1]\) where
  \(\mathfrak{Y}\) is the set of all possible measurement sets and
  \(\hat{\mathfrak{Y}}\) is the set of all possible true measurable
  values.
\item
  A deterministic process model that describes the measured quantities
  in terms of unknown, possibly unobserved parameters \(\theta\), i.e.~a
  function \(d: \mathfrak{theta} -> \hat{mathfrak{Y}}\).
\item
  A prior model that probabilistically describes how likely any possible
  set of parameter values is, without considering any information
  included in the measurement model, i.e.~a probability density function
  \(\pi : \mathfrak{theta} \rightarrow [0, 1]\).
\end{itemize}

Together these components determine a joint probability function that
encapsulates the Bayesian statistical model, i.e.~a function
\(p: \mathfrak{Y}\times\mathfrak{theta}\rightarrow[0,1]\) such that for
any \(\theta\) and \(y\), \(p(\theta, y) = \pi(\theta)l(y\mid\theta)\).
Substantive questions about the implications of the assumptions implicit
in the model components can be formulated in terms of this joint density
function.

In particular, questions about what the model implies, given a
particular measurement set \(y\) can be formulated in terms of the
conditional probability density function
\(p_{\mid y}:\mathfrak{theta}\rightarrow[0,1]\). Bayes's theorem
guarantees that this conditional density is proportional to the product
of the prior and likelihood,
i.e.~\(p_{\mid y}(\theta) \propto \pi(\theta)l(y\mid\theta)\) See
\citep[Ch 1]{gelmanBayesianDataAnalysis2020a} for an extended discussion
of the theory of Bayesian statistical inference.

The principal computational challenge for Bayesian statistical inference
is to approximate the values of integrals of the joint density function
conditional on a known measurement assignment \(y\), also known as the
posterior distribution. For most non-trivial statistical models the
posterior distribution cannot be integrated analytically, so numerical
approximation is required. Markov Chain Monte Carlo or ``MCMC'' is a
popular method for addressing this problem that uses a Markov chain with
known properties to generate samples from a target probability
distribution that can be used to perform Monte Carlo integration.
Hamiltonian Monte Carlo makes it possible, given an appropriate choice
of hyperparameters, to efficiently generate samples from even a high
dimensional continuous posterior distribution, by calculating the
gradients of the log-scale joint density function. Adaptive Hamiltonian
Monte Carlo, as used by Stan, uses a range of well-tested methods to
tune these hyperparameters, allowing efficient MCMC for high dimensional
posterior distributions with minimal user input.

Alternatives to MCMC for numerically approximating integrals over
posterior distributions include grid sampling \citep[ Ch.
10]{gelmanBayesianDataAnalysis2020a}, rejection sampling \citep[Ch.
10]{gelmanBayesianDataAnalysis2020a}, importance sampling
\citep{vehtariParetoSmoothedImportance2022} and distributional
approaches including variational inference and Laplace approximation
\citep[Ch. 13]{gelmanBayesianDataAnalysis2020a}.

The Laplace approximation is of particular interest because, as
mentioned above, this has been used for approximate Bayesian inference
in a similar context to Maud. Laplace approximation of a posterior
distribution works by first finding the mode of the posterior
distribution, i.e.~the `maximum a posteriori' parameter configuration
with the highest posterior density. Next, the second derivative of the
posterior distribution at this point is found and used to generate a
normal distribution that either approximates the target distribution, or
can in turn be used to generate such an approximation.

Maud employs adaptive Hamiltonian Monte Carlo to perform posterior
inference for a Bayesian statistical model where the process model is
the kinetic model described above. The next two subsections describe
Maud's prior and measurement models.

### 2024-01-12 paper/supporting_information.qmd
review response changes

>>>>

# Multimodal posterior distributions

While we have not yet observed this in practice, we expect that Maud is capable
of accurately sampling mildly multi-modal posterior distributions, i.e. those
for which regions of parameter space with high probability density are not
very sharply separated. This is because this ability depends primarily on the
underlying inference algorithm, and adaptive Hamiltonian Monte Carlo is known
to be able to sample many such posterior distributions. See, for example,
@mangiolaSccompRobustDifferential2023, which reports the successful use of
adaptive Hamiltonian Monte Carlo to sample posterior distributions involving
mixtures of Gaussian distributions, which are typically multi-modal.

Moreover, in cases where it fails to sample a multi-modal posterior distribution
adaptive Hamiltonian Monte Carlo typically exhibits poor mixing and divergent
transitions, which Maud is set up to detect automatically, making this case easy
for users to diagnose.

Nonetheless, it is also well known that strongly multi-modal posterior
distributions pose a problem for adaptive Hamiltonian Monte Carlo, as well as
other MCMC algorithms. See [@betancourtmichaelMarkovChainMonte2020, §3.2] for
a general discussion of this issue. @mangoubiDoesHamiltonianMonte2018 explores
the related question of the relative performance of Hamiltonian Monte Carlo vs
random walk Metropolis Hastings, finding that, for certain cases where the target
distribution exhibits highly disconnected and similarly shaped modes, the two
algorithms have similar performance.

It is therefore important to pay careful attention to MCMC diagnostics and
monitor developments in computational statistics that might expand the range of
posterior distributions that can practically be sampled.

@zamora-silleroEfficientCharacterizationHighdimensional2011 introduces
the method HYPERSPACE, which is designed to identify disconnected regions
of a biochemical parameter space and characterise them by estimating
their dimensions. HYPERSPACE does not perform parameter inference and is
therefore not directly comparable with algorithms like adaptive Hamiltonian
Monte Carlo as used in Maud, but could plausibly be adapted into such an
algorithm, for example by supplementation with importance sampling (see
@vehtariParetoSmoothedImportance2022). However, any such algorithm would
be limited by HYPERSPACE's dependency on a random walk Metropolis Hastings
algorithm. This kind of algorithm's performance is known to scale poorly
compared with Hamiltonian Monte Carlo algorithms as the number of parameters
increases: see @mangoubiDoesHamiltonianMonte2018. Since Maud aims to fit models
with hundreds of parameters, this means that it is unlikely that HYPERSPACE can
directly be used to improve Maud's efficiency.

### 2024-01-12 paper/supporting_information.tex
review response changes

\title[]{Supporting information - Bayesian regression facilitates
quantitative modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi

@@

>>>>

\title[]{Supporting information - Bayesian regression facilitates
quantitative modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\hypertarget{multimodal-posterior-distributions}{%
\section{Multimodal posterior
distributions}\label{multimodal-posterior-distributions}}

While we have not yet observed this in practice, we expect that Maud is
capable of accurately sampling mildly multi-modal posterior
distributions, i.e.~those for which regions of parameter space with high
probability density are not very sharply separated. This is because this
ability depends primarily on the underlying inference algorithm, and
adaptive Hamiltonian Monte Carlo is known to be able to sample many such
posterior distributions. See, for example,
\citet{mangiolaSccompRobustDifferential2023}, which reports the
successful use of adaptive Hamiltonian Monte Carlo to sample posterior
distributions involving mixtures of Gaussian distributions, which are
typically multi-modal.

Moreover, in cases where it fails to sample a multi-modal posterior
distribution adaptive Hamiltonian Monte Carlo typically exhibits poor
mixing and divergent transitions, which Maud is set up to detect
automatically, making this case easy for users to diagnose.

Nonetheless, it is also well known that strongly multi-modal posterior
distributions pose a problem for adaptive Hamiltonian Monte Carlo, as
well as other MCMC algorithms. See
\citep[§3.2]{betancourtmichaelMarkovChainMonte2020} for a general
discussion of this issue. \citet{mangoubiDoesHamiltonianMonte2018}
explores the related question of the relative performance of Hamiltonian
Monte Carlo vs random walk Metropolis Hastings, finding that, for
certain cases where the target distribution exhibits highly disconnected
and similarly shaped modes, the two algorithms have similar performance.

It is therefore important to pay careful attention to MCMC diagnostics
and monitor developments in computational statistics that might expand
the range of posterior distributions that can practically be sampled.

\citet{zamora-silleroEfficientCharacterizationHighdimensional2011}
introduces the method HYPERSPACE, which is designed to identify
disconnected regions of a biochemical parameter space and characterise
them by estimating their dimensions. HYPERSPACE does not perform
parameter inference and is therefore not directly comparable with
algorithms like adaptive Hamiltonian Monte Carlo as used in Maud, but
could plausibly be adapted into such an algorithm, for example by
supplementation with importance sampling (see
\citet{vehtariParetoSmoothedImportance2022}). However, any such
algorithm would be limited by HYPERSPACE's dependency on a random walk
Metropolis Hastings algorithm. This kind of algorithm's performance is
known to scale poorly compared with Hamiltonian Monte Carlo algorithms
as the number of parameters increases: see
\citet{mangoubiDoesHamiltonianMonte2018}. Since Maud aims to fit models
with hundreds of parameters, this means that it is unlikely that
HYPERSPACE can directly be used to improve Maud's efficiency.
